Temperature and wave exposure as drivers of morphology in two kelp species

### 2019-07-30 morphology_paper.Rmd
refining outputs

---
title: Temperature and wave exposure as drivers of morphology in two kelp species
  along the coast of South Africa
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
geometry: margin=1in
output:
  html_notebook: default
  pdf_document:
    fig_caption: yes
    fig_crop: no
    fig_height: 7
    fig_width: 7
  html_document:
    df_print: paged
  word_document:
    reference_docx: frontiers_template.docx
papersize: A4
tables: yes
fontsize: 10pt
---

@@

```{r Temperature data prep, message=FALSE, warning=FALSE, include=FALSE}
## Data wrangling and preparation
## Preparing data
# Complete lon/lats
kelp_sites <- read.csv("metadata/site_meta_complete.csv", sep = ";")

@@

# Laminaria
lam_data <- read.csv("data/morph_data/lam_morph_data.csv", sep = ";") %>%
  mutate(site = replace(site, site == "Buffels Bay", "Buffelsbaai")) %>%
  # NB: Even though I am changing the name, there will still be no lon/lat values for this site
  # We need to decide how we want to handle this. Give it the same lon/lat as Buffelsbaai?
  mutate(site = replace(site, site == "Buffels Bay South", "Buffelsbaai South")) %>%
  # NB: This may also not be an appropriate change
  mutate(site = replace(site, site == "Bordjies reef North", "Bordjies Deep")) %>%
  # Not using Rocky Bank in this study
  filter(site != "Rocky Bank")

@@

# The patched Roman Rock temperatures
load("data/rr_patch.Rdata")

## Subset temperatures for collection sites
temps <- SACTNmonthly_v4.1 %>% 
  filter(site %in% eck_data$site | site %in% lam_data$site) %>% 
  select(site, date, temp) %>% 
  rbind(., interp_sites) %>% 
  filter(site!= "Roman Rock") %>% 
  rbind(., rr_patch)

## Temperature analyses
# Up next we need to calculate the relevant temperature statistics for each site. These are: #mean, range, and sd, for Annual, February, and August climatologies.
# Monthly clims

temp_feb_aug <- temps %>% 
  mutate(date = lubridate::month(date, label = TRUE)) %>% 
  group_by(site, date) %>% 
  summarise(mean_temp = mean(temp, na.rm = T),
            min_temp = min(temp, na.rm = T),
            max_temp = max(temp, na.rm = T),
            range_temp = range(temp, na.rm = T)[2]-range(temp, na.rm = T)[1],
            sd_temp = sd(temp, na.rm = T)) %>%
  filter(date %in% c("Feb", "Aug")) %>% 
  ungroup()

# Annual clims
temp_ann <- temps %>% 
  mutate(date = lubridate::month(date, label = TRUE)) %>%
  group_by(site, date) %>% 
  summarise(mean_temp = mean(temp, na.rm = T),
            min_temp = min(temp, na.rm = T),
            max_temp = max(temp, na.rm = T),
            range_temp = range(temp, na.rm = T)[2]-range(temp, na.rm = T)[1],
            sd_temp = sd(temp, na.rm = T)) %>% 
  ungroup() %>% 
  select(-date) %>% 
  group_by(site) %>% 
  summarise_all(funs(mean(., na.rm = T))) %>% 
  mutate(date = "Ann") %>% 
  select(site, date, everything())

# Combine
temp_clim <- rbind(temp_feb_aug, temp_ann) %>% 
  group_by(site) %>% 
  gather(key = statistic, value = value, 
         mean_temp:sd_temp) %>% 
  mutate(clim_stat = paste(date, statistic, sep = "_")) %>% 
  select(-date, -statistic) %>% 
  spread(key = clim_stat, value = value)
```

```{r Temperature variable processing, message=FALSE, warning=FALSE, include=FALSE}
### Plots for temperature variables 
positions_temp <- c("Betty's Bay", "Roman Rock","Kalk Bay", "A-Frame", "Miller's Point", "Baboon Rock",  "Batsata Rock", "Buffelsbaai","Buffelsbaai South","Olifantsbos","Scarborough", "Soetwater", "Kommetjie","Hout Bay","Oudekraal","Sea Point","De Hoop", "St Helena Bay")

# Annual temp
temp_plot <- ggplot(data = temp_ann, aes(x = site, y = mean_temp)) +
  geom_point() + 
  scale_x_discrete(limits = positions_temp) + 
  ylim(8, 22) + 
  geom_errorbar(aes(x = site, ymin = mean_temp - sd_temp, ymax = mean_temp + sd_temp)) + 
  geom_point(data = temp_ann, aes(x = site, y = max_temp, colour = "blue")) + 
  geom_point(data = temp_ann, aes(x = site, y = min_temp, colour = "pink")) + 
  xlab("") + 
  ylab("") + 
  theme(plot.title = element_text(hjust = 0)) +
  theme(legend.position = "none") + 
  coord_flip() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank())

@@

# Calculate the necessary stats
# Monthly clims
wave_feb_aug <- wave_model %>% 
  mutate(date = lubridate::month(date, label = TRUE)) %>% 
  filter(site != "FB025") %>% # FB_25 has issues with 'hs' and 'tp'
  group_by(site, date) %>% 
  summarise_all(funs(mean = mean, sd = sd, min = min, max = max), na.rm = T) %>%
  filter(date %in% c("Feb", "Aug")) %>% 
  ungroup()

@@

```{r wave variable processing, warning=FALSE, include=FALSE, echo = FALSE, message=FALSE}
### Plots for wave variables
#Annual wave plots
pos_temp <- c("Cool temperate", "Warm temperate")

wave_plot_A <- ggplot(data = wave_ann, aes(x = site, y = hs_mean)) +
  geom_point() + 
  scale_x_discrete(limits = positions_temp) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5)) +
  geom_errorbar(aes(x = site, ymin = hs_mean - hs_sd, ymax = hs_mean + hs_sd)) + 
  xlab("") + 
  ylab("") + 
  theme(legend.position = "none") + 
  theme(plot.title = element_text(hjust = 0)) + 
  coord_flip() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank())

wave_plot_B <- ggplot(data = wave_ann, aes(x = site, y = tp_mean)) +
  geom_point() + 
  scale_x_discrete(limits = positions_temp) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5)) +
  geom_errorbar(aes(x = site, ymin = tp_mean - tp_sd, ymax = tp_mean + tp_sd)) + 
  xlab("") + 
  ylab("") + 
  theme(legend.position = "none") + 
  theme(plot.title = element_text(hjust = 0)) + 
  coord_flip() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank())  
```

@@

```{r deep ecklonia processing, message=FALSE, warning=FALSE, include=FALSE}
## Preparing data
# Ecklonia
eck_data <- read.csv("data/morph_data/eck_morph_data.csv", sep = ";") %>%
  filter(site!= "Rocky Bank") %>%   # Not using Rocky Bank in this study
  filter(site!= "De Hoop")  
# group_by(wave_exposure, frond_mass) %>% 
# summarise(mean = mean(frond_mass:total_length, na.rm = TRUE)) %>% 
# ungroup()
eck_deep <- eck_data %>% 
  filter(population == "Deep")
eck_shall <- eck_data %>% 
  filter(population == "Shallow")

@@

```{r shallow ecklonia temp plots, echo=FALSE, message=FALSE, warning=FALSE, fig.cap ='Boxplot of morphological characteristics for shallow population *E. maxima* grouped by temperature regime and wave exposure.'}
ggarrange(EST1,EST2,EST3,EST4,EST5,EST6,EST7,EST8,EST9,EST10, ESW1,ESW2,ESW3,ESW4,ESW5,ESW6,ESW7,ESW8, ESW9, labels = "AUTO", common.legend = FALSE, legend = "bottom")
```

@@

#calculate mean values for lamanaria morphology
lam_mean <- lam_df %>% 
  subset(select = -c(1,2)) %>% 
  group_by(site) %>% 
  summarise_all(funs(mean), na.rm = T) %>% 
  ungroup()

@@

#remove Bordjies North from data-frame
lam_mean <- lam_mean[-c(5),]

lam_meta <- kelp_sites %>% 
  filter(population == "Deep") %>% 
  filter(species == "Both")

lam_meta <- merge(lam_sites, kelp_sites, by = "site")
#drop useless columns
lam_meta <- lam_meta[-c(1,9,11), ]

@@

```{r rda plot lamanaria, echo=FALSE, message=FALSE, warning=FALSE}
# Plot RDA
ggord(rda_lam, lam_meta$region, ellipse = FALSE, hull = FALSE, grp_title = "Region", ptslab = TRUE, parse = TRUE, addsize = 3.5, ext = 0.9)  
```

```{r rda ecklonia processing, echo=FALSE, message=FALSE, warning=FALSE}

#filter to select only deep population ecklonia
eck_df <- eck_data %>% 
  filter(population == "Deep")

@@

#remove St Helena Bay for now as wave model not available for that site
eck_mean <- eck_mean[-c(13), ]

@@

#remove Miller's Point site as it is causing issues, also this keeps the amount of western and eastern sites more balanced
eck_sites <- eck_sites %>% 
  filter(site != "Miller's Point")

eck_meta <- read.csv("metadata/eck_meta.csv", sep = ";") 
eck_meta <- eck_meta %>% 
  filter(population == "Deep") %>% 
  mutate(site = replace(site, site == "Miller's Point", "Miller's Point"))

eck_meta <- merge(eck_meta, eck_sites,  by = "site")

#standardise all abiotic measurements
bio_eck<- eck_mean %>%
  filter(site != "Miller's Point") %>% 
  subset(select = -c(1)) %>%
  decostand(method = "standardize")

@@

# standardise all abiotic measurements
abio_eck <- abio_eck %>%
  filter(site != "Miller's Point") %>% 
  subset(select = -c(1)) %>% 
  decostand(method = "standardize")

@@

step.forward <- ordistep(rda(bio_eck ~ 1, data = abio_eck),
                         scope=formula(rda_all_eck), direction = "forward", pstep = 1000)
#calculate variance inlfation factor
vif.cca(rda_all_eck)

@@

# Plot RDA
ggord(rda_eck, eck_meta$region, ellipse = FALSE, hull = FALSE, grp_title = "Sites", ptslab = TRUE, parse = TRUE, addsize = 3.5, ext = 0.95)  # looking at the raw code, this is plotting the 'wa scores', the blue 
```

@@

#filter to select only deep population ecklonia and remove Millers B and C for now...
eck_df_sh <- eck_data %>% 
  filter(population == "Shallow") %>% 
  filter(site != "Miller's Point B") %>% 
  filter(site != "Miller's Point C")

@@

#calculate mean values for lamanaria morphology
eck_mean_sh <- eck_df_sh %>% 
  subset(select = -c(1,2)) %>% 
  group_by(site) %>% 
  summarise_all(funs(mean), na.rm = T) %>% 
  ungroup()

@@

eck_meta_sh <- read.csv("metadata/eck_meta.csv", sep = ";") 
eck_meta_sh <- eck_meta_sh %>% 
  filter(population == "Shallow")

eck_meta_sh <- merge(eck_meta_sh, eck_sites_sh,  by = "site")

@@

#use site eck_site to align appropriate data from abio_clim data from the lamanaria processing section
abio_eck_sh <- merge(eck_sites_sh, abio_clim, by = "site")

@@

#run RDA for abiotic variables
rda_all_eck <- rda(bio_eck_sh ~ ., data = abio_eck_sh)

@@

step.forward <- ordistep(rda(bio_eck ~ 1, data = abio_eck),
                         scope=formula(rda_all_eck), direction = "forward", pstep = 1000)
#calculate variance inlfation factor
vif.cca(rda_all_eck)

# Plot RDA
ggord(rda_eck, eck_meta$region, ellipse = FALSE, hull = FALSE, grp_title = "Sites", ptslab = TRUE, parse = TRUE, addsize = 3.5, ext = 0.95)  # looking at the raw code, this is plotting the 'wa scores', the blue

@@

>>>>

---
title: Temperature and wave exposure as drivers of morphology in two kelp species
  along the coast of South Africa
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
geometry: margin=1in
output:
  html_notebook: 
    css: frontiers.csl
    fig_caption: yes
    fig_height: 7
    fig_width: 7
  pdf_document:
    fig_caption: yes
    fig_crop: no
    fig_height: 7
    fig_width: 7
  html_document:
    df_print: paged
  word_document:
    reference_docx: frontiers_template.docx
papersize: A4
tables: yes
fontsize: 10pt
---

@@

```{r Temperature data prep, message=FALSE, warning=FALSE, include=FALSE}
## Data wrangling and preparation
## Preparing data
# Complete lon/lats

@@

# Laminaria
lam_data <- read.csv("data/morph_data/lam_morph_data.csv", sep = ";") %>%
  mutate(site = replace(site, site == "Buffels Bay", "Buffelsbaai")) %>%
  # NB: Even though I am changing the name, there will still be no lon/lat values for this site
  # We need to decide how we want to handle this. Give it the same lon/lat as Buffelsbaai?
  mutate(site = replace(site, site == "Buffels Bay South", "Buffelsbaai South")) %>%
  # NB: This may also not be an appropriate change
  mutate(site = replace(site, site == "Bordjies North", "Bordjies")) %>%
  # Not using Rocky Bank in this study
  filter(site != "Rocky Bank")

@@

# The patched Roman Rock temperatures
load("data/rr_patch.Rdata")
kelp_sites <- read.csv("metadata/site_meta_complete.csv", sep = ";") %>% 
  mutate(site = replace(site, site == "Bakoven", "Sea Point"))

## Subset temperatures for collection sites
temps <- SACTNmonthly_v4.1 %>% 
  filter(site %in% eck_data$site | site %in% lam_data$site) %>% 
  select(site, date, temp) %>% 
  rbind(., interp_sites) %>% 
  filter(site!= "Roman Rock") %>% 
  rbind(., rr_patch)
## Temperature analyses
# Up next we need to calculate the relevant temperature statistics for each site. These are: #mean, range, and sd, for Annual climatology.
# Annual clims
temp_ann <- temps %>% 
  mutate(date = lubridate::month(date, label = TRUE)) %>%
  group_by(site, date) %>% 
  summarise(mean_temp = mean(temp, na.rm = T),
            min_temp = min(temp, na.rm = T),
            max_temp = max(temp, na.rm = T),
            range_temp = range(temp, na.rm = T)[2]-range(temp, na.rm = T)[1],
            sd_temp = sd(temp, na.rm = T)) %>% 
  ungroup() %>% 
  select(-date) %>% 
  group_by(site) %>% 
  summarise_all(funs(mean(., na.rm = T))) %>% 
  mutate(date = "Ann") %>% 
  select(site, date, everything())
```

```{r Temperature variable processing, message=FALSE, warning=FALSE, include=FALSE}
### Plots for temperature variables 
positions_temp <- c("Betty's Bay", "Roman Rock","Kalk Bay", "A-Frame", "Miller's Point", "Baboon Rock",  "Batsata Rock", "Buffelsbaai","Buffelsbaai South","Olifantsbos","Scarborough", "Soetwater", "Kommetjie","Hout Bay","Oudekraal","Sea Point","St Helena Bay")

# Annual temp
temp_plot <- ggplot(data = temp_ann, aes(x = site, y = mean_temp)) +
  geom_point() + 
  scale_x_discrete(limits = positions_temp) + 
  ylim(9, 21) + 
  geom_errorbar(aes(x = site, ymin = mean_temp - sd_temp, ymax = mean_temp + sd_temp)) + 
  geom_point(data = temp_ann, aes(x = site, y = max_temp, colour = "blue")) + 
  geom_point(data = temp_ann, aes(x = site, y = min_temp, colour = "pink")) + 
  xlab("") + 
  ylab("") + 
  theme(plot.title = element_text(hjust = 0)) +
  theme(legend.position = "none") + 
  coord_flip() 
```

@@

@@

```{r wave variable processing, warning=FALSE, include=FALSE, echo = FALSE, message=FALSE}
### Plots for wave variables
#Annual wave plots

positions_wave <- c("Betty's Bay", "Roman Rock","Kalk Bay", "A-Frame", "Miller's Point", "Baboon Rock",  "Batsata Rock","Bordjies Deep","Buffelsbaai","Buffelsbaai South","Olifantsbos","Scarborough", "Soetwater", "Kommetjie","Hout Bay","Oudekraal","Sea Point","St. Helena Bay")

wave_plot_A <- ggplot(data = wave_ann, aes(x = site, y = hs_mean)) +
  geom_point() + 
  scale_x_discrete(limits = positions_wave) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5)) +
  geom_errorbar(aes(x = site, ymin = hs_mean - hs_sd, ymax = hs_mean + hs_sd)) + 
  xlab("") + 
  ylab("") + 
  theme(legend.position = "none") + 
  theme(plot.title = element_text(hjust = 0)) + 
  coord_flip()

wave_plot_B <- ggplot(data = wave_ann, aes(x = site, y = tp_mean)) +
  geom_point() + 
  scale_x_discrete(limits = positions_wave) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5)) +
  geom_errorbar(aes(x = site, ymin = tp_mean - tp_sd, ymax = tp_mean + tp_sd)) + 
  xlab("") + 
  ylab("") + 
  theme(legend.position = "none") + 
  theme(plot.title = element_text(hjust = 0)) + 
  coord_flip() 
```

@@

```{r deep ecklonia processing, message=FALSE, warning=FALSE, include=FALSE}
## Preparing data
# Ecklonia
eck_data <- read.csv("data/morph_data/eck_morph_data.csv", sep = ";") %>%
  filter(site!= "Rocky Bank") %>%   # Not using Rocky Bank in this study
  filter(site!= "De Hoop")  
# group_by(wave_exposure, frond_mass) %>% 
# summarise(mean = mean(frond_mass:total_length, na.rm = TRUE)) %>% 
# ungroup()
eck_deep <- eck_data %>% 
  filter(population == "Deep")
eck_shall <- eck_data %>% 
  filter(population == "Shallow") %>% 
  mutate(site = replace(site, site == "Bakoven", "Sea Point"))

@@

```{r shallow ecklonia temp plots, echo=FALSE, message=FALSE, warning=FALSE, fig.cap ='Boxplot of morphological characteristics for shallow population *E. maxima* grouped by temperature regime and wave exposure.'}
ggarrange(EST1,EST2,EST3,EST4,EST5,EST6,EST7,EST8,EST9,EST10,ESW1,ESW2,ESW3,ESW4,ESW5,ESW6,ESW7,ESW8, ESW9, labels = "AUTO", common.legend = FALSE, legend = "bottom")
```

@@

#calculate mean values for lamanaria morphology
lam_mean <- lam_df %>% 
  subset(select = -c(1,2)) %>% 
  group_by(site) %>% 
  summarise_all(funs(mean), na.rm = T) %>% 
  filter(site != "Roman Rock") %>% 
  ungroup()

@@

@@

lam_meta_rda <- merge(lam_sites, kelp_sites, by = "site")
lam_meta_rda <- lam_meta_rda[-c(1,9,11,14), ]

# Plot RDA
lam_rda <- ggord(rda_lam, lam_meta_rda$region, ellipse = TRUE, hull = FALSE, grp_title = "Region", ptslab = TRUE, parse = TRUE, addsize = 3.5, ext = 0.9) 
```

```{r rda deep ecklonia processing, echo=FALSE, message=FALSE, warning=FALSE}

#filter to select only deep population ecklonia
eck_df <- eck_data %>% 
  filter(population == "Deep") %>% 
  filter(site != "Miller's Point")

@@

#remove St Helena Bay for now as wave model not available for that site
eck_mean <- eck_mean %>% 
  filter(site != "St Helena Bay")

@@

#standardise all abiotic measurements
bio_eck<- eck_mean %>%
  subset(select = -c(1)) %>%
  decostand(method = "standardize")

@@

# standardise all abiotic measurements
abio_eck <- abio_eck %>%
  subset(select = -c(1)) %>% 
  decostand(method = "standardize")

@@

step.forward <- ordistep(rda(bio_eck ~ 1, data = abio_eck),
                         scope=formula(rda_all_eck), direction = "forward", pstep = 1000)
#calculate variance inlfation factor
#vif.cca(rda_all_eck)

@@

eck_meta_rda <- merge(eck_sites, kelp_sites, by = "site")
eck_meta_rda <- eck_meta_rda %>% 
  filter(population == "Deep")

# Plot RDA
rda_eck_deep <- ggord(rda_eck, ellipse = TRUE, hull = FALSE, ptslab = TRUE, parse = TRUE, addsize = 3.5, ext = 0.95)  # looking at the raw code, this is plotting the 'wa scores', the blue 
```

@@

#filter to select only deep population ecklonia and remove Millers B and C for now...
eck_df_sh <- eck_data %>%
  filter(population == "Shallow") %>% 
  mutate(site = replace(site, site == "Miller's Point A", "Miller's Point")) %>% 
  mutate(site = replace(site, site == "Bakoven", "Sea Point"))

@@

#calculate mean values for shallow ecklonia morphology
eck_mean_sh <- eck_df_sh %>% 
  subset(select = -c(1,2)) %>% 
  group_by(site) %>% 
  summarise_all(funs(mean), na.rm = T) %>% 
  ungroup()

@@

@@

#use site eck_site to align appropriate data from abio_clim data from the lamanaria processing section
abio_eck_sh <- merge(abio_clim, eck_sites_sh, by = "site")

@@

#run RDA for abiotic variables
rda_all_ecksh <- rda(bio_eck_sh ~ ., data = abio_eck_sh)

#summary of rda output
summary(rda_all_ecksh)

@@

step.forward <- ordistep(rda(bio_eck_sh ~ 1, data = abio_eck_sh),
                         scope=formula(rda_all_eck), direction = "forward", pstep = 1000)
#calculate variance inlfation factor
#vif.cca(rda_all_eck)

#use pearson correlations to determine best abiotic variables for model
cor_eck_abio <- round(cor(abio_eck_sh, use = "pair"), 2)
cor_eck_abio

# Parsimonious RDA
rda_eck_shallow <- rda(bio_eck_sh ~ hs_max + hs_mean + sd_temp + mean_temp, data = abio_eck_sh)
summary(rda_eck_shallow)

#visual screenshot of rda axes to help determine which axes explain the most variation
screeplot(rda_eck_shallow)

#retreive adjusted R2 value which measures the unbiased amount of explained variation
R2adj <- RsquareAdj(rda_eck_shallow)$adj.r.squared
R2adj

# Plot RDA
rda_eck_shallow <- ggord(rda_eck_shallow, eck_meta_sh$region,ellipse = TRUE, hull = FALSE, grp_title = "Sites", ptslab = TRUE, parse = TRUE, addsize = 3.5, ext = 0.9, coord_fix = TRUE)  # looking at the raw code, this is plotting the 'wa scores', the blue

@@

```{r rda plots, echo=FALSE}
ggarrange(lam_rda, rda_eck_deep,rda_eck_shallow, common.legend = TRUE, labels = "AUTO", legend = "bottom")
```
